---
title: "YFinance"
author: "Mkshots"
format: html
editor: visual
jupyter: python3
---

## Phase 1: Workbook set-up & YFinance Downloads

```{r}
library(reticulate)
py_config()
```

```{r}
library(reticulate)
py_install(c("yfinance", "pandas", "openpyxl", "datetime", "pytz","IPython", "scipy"), pip = TRUE)
```

```{python}
import yfinance as yf
import pandas as pd
import datetime as datetime
import pytz
from IPython.display import display
```

Path

```{python}
base_dir = r"your path location here"
```

For CSVs like a data source:

```{python}
import os
import re
from datetime import datetime

import pandas as pd
import yfinance as yf

# =====================================================
# SETTINGS
# =====================================================

tickers = [
    "TICKER 1",
    "TICKER 2",
    "TICKER 3",
    "TICKER 4",
    "TICKER 5",
#add more tickers as required.
]

start_year = 2020

# base_dir should already be defined in a previous chunk
# Example:
# base_dir = r"C:\Users\username\path"

today = datetime.today().strftime("%Y%m%d")

# =====================================================
# CREATE FOLDERS
# =====================================================

income_dir = os.path.join(base_dir, "income_statements")
balance_dir = os.path.join(base_dir, "balance_sheets")
cashflow_dir = os.path.join(base_dir, "cashflows")
info_dir = os.path.join(base_dir, "company_info")
price_dir = os.path.join(base_dir, "price_history")

for folder in [
    income_dir,
    balance_dir,
    cashflow_dir,
    info_dir,
    price_dir
]:
    os.makedirs(folder, exist_ok=True)

# =====================================================
# HELPERS
# =====================================================

def safe_filename(name):
    name = str(name).strip()
    name = re.sub(r'[\\/*?:"<>|]', "", name)
    name = re.sub(r"\s+", "_", name)
    return name[:100]

def get_company_name(ticker_obj, fallback):

    try:

        info = ticker_obj.info

        company = (
            info.get("longName")
            or info.get("shortName")
            or fallback
        )

        return safe_filename(company)

    except Exception:

        return safe_filename(fallback)

def prepare_statement(df):

    if df is None:
        return pd.DataFrame()

    if df.empty:
        return pd.DataFrame()

    cols = pd.to_datetime(df.columns, errors="coerce")

    valid = cols.notna()

    df = df.loc[:, valid]
    cols = cols[valid]

    keep = cols.year >= start_year

    df = df.loc[:, keep]
    cols = cols[keep]

    if df.empty:
        return pd.DataFrame()

    df.columns = cols.strftime("%Y-%m-%d")

    return df

def convert_to_long(df, ticker, company):

    output = (
        df.reset_index()
          .rename(columns={"index": "line_item"})
          .melt(
              id_vars="line_item",
              var_name="fiscal_date",
              value_name="value"
          )
    )

    output["ticker"] = ticker
    output["company_name"] = company

    return output

# =====================================================
# STORAGE
# =====================================================

all_income = []
all_balance = []
all_cashflow = []
all_info = []
all_prices = []

# =====================================================
# MAIN LOOP
# =====================================================

for tk in tickers:

    print(f"Processing {tk}")

    try:

        ticker_obj = yf.Ticker(tk)

        company = get_company_name(
            ticker_obj,
            tk
        )

        # ==========================================
        # INCOME STATEMENT
        # ==========================================

        income = prepare_statement(
            ticker_obj.financials
        )

        if not income.empty:

            file_name = (
                f"{today}_{tk}_{company}_income_statement.csv"
            )

            income.to_csv(
                os.path.join(
                    income_dir,
                    file_name
                )
            )

            all_income.append(
                convert_to_long(
                    income,
                    tk,
                    company
                )
            )

        # ==========================================
        # BALANCE SHEET
        # ==========================================

        balance = prepare_statement(
            ticker_obj.balance_sheet
        )

        if not balance.empty:

            file_name = (
                f"{today}_{tk}_{company}_balance_sheet.csv"
            )

            balance.to_csv(
                os.path.join(
                    balance_dir,
                    file_name
                )
            )

            all_balance.append(
                convert_to_long(
                    balance,
                    tk,
                    company
                )
            )

        # ==========================================
        # CASH FLOW
        # ==========================================

        cashflow = prepare_statement(
            ticker_obj.cashflow
        )

        if not cashflow.empty:

            file_name = (
                f"{today}_{tk}_{company}_cashflow.csv"
            )

            cashflow.to_csv(
                os.path.join(
                    cashflow_dir,
                    file_name
                )
            )

            all_cashflow.append(
                convert_to_long(
                    cashflow,
                    tk,
                    company
                )
            )

        # ==========================================
        # COMPANY INFO
        # ==========================================

        try:

            info = ticker_obj.info

            info_df = pd.DataFrame([{
                "ticker": tk,
                "company_name": company,

                "marketCap": info.get("marketCap"),
                "enterpriseValue": info.get("enterpriseValue"),
                "sharesOutstanding": info.get("sharesOutstanding"),
                "floatShares": info.get("floatShares"),

                "beta": info.get("beta"),

                "debtToEquity": info.get("debtToEquity"),
                "currentRatio": info.get("currentRatio"),
                "quickRatio": info.get("quickRatio"),

                "returnOnAssets": info.get("returnOnAssets"),
                "returnOnEquity": info.get("returnOnEquity"),

                "operatingMargins": info.get("operatingMargins"),
                "profitMargins": info.get("profitMargins"),

                "sector": info.get("sector"),
                "industry": info.get("industry"),

                "country": info.get("country"),
                "currency": info.get("currency"),

                "exchange": info.get("exchange")
            }])

            info_file = (
                f"{today}_{tk}_{company}_company_info.csv"
            )

            info_df.to_csv(
                os.path.join(
                    info_dir,
                    info_file
                ),
                index=False
            )

            all_info.append(info_df)

        except Exception as e:

            print(
                f"Company info error for {tk}: {e}"
            )

        # ==========================================
        # PRICE HISTORY
        # ==========================================

        try:

            prices = ticker_obj.history(
                period="5y",
                auto_adjust=True
            )

            if not prices.empty:

                prices = prices.reset_index()

                prices["ticker"] = tk
                prices["company_name"] = company

                price_file = (
                    f"{today}_{tk}_{company}_price_history.csv"
                )

                prices.to_csv(
                    os.path.join(
                        price_dir,
                        price_file
                    ),
                    index=False
                )

                all_prices.append(prices)

        except Exception as e:

            print(
                f"Price history error for {tk}: {e}"
            )

    except Exception as e:

        print(
            f"Error processing {tk}: {e}"
        )

# =====================================================
# COMBINED FILES
# =====================================================

if all_income:

    pd.concat(
        all_income,
        ignore_index=True
    ).to_csv(
        os.path.join(
            base_dir,
            f"{today}_ALL_income_statements.csv"
        ),
        index=False
    )

if all_balance:

    pd.concat(
        all_balance,
        ignore_index=True
    ).to_csv(
        os.path.join(
            base_dir,
            f"{today}_ALL_balance_sheets.csv"
        ),
        index=False
    )

if all_cashflow:

    pd.concat(
        all_cashflow,
        ignore_index=True
    ).to_csv(
        os.path.join(
            base_dir,
            f"{today}_ALL_cashflows.csv"
        ),
        index=False
    )

if all_info:

    pd.concat(
        all_info,
        ignore_index=True
    ).to_csv(
        os.path.join(
            base_dir,
            f"{today}_ALL_company_info.csv"
        ),
        index=False
    )

if all_prices:

    pd.concat(
        all_prices,
        ignore_index=True
    ).to_csv(
        os.path.join(
            base_dir,
            f"{today}_ALL_price_history.csv"
        ),
        index=False
    )

print("Download complete.")
```

Getting the Risk-Free Rate

```{python}
# =====================================================
# DOWNLOAD RISK-FREE RATE (^TNX)
# =====================================================

import yfinance as yf
import pandas as pd
import os
from datetime import datetime

today = datetime.today().strftime("%Y%m%d")

print("Downloading US 10Y Treasury Yield...")

# Yahoo ticker for US 10Y Treasury Yield
tnx = yf.download(
    "^TNX",
    start="2020-01-01",
    auto_adjust=True,
    progress=False
)

# =====================================================
# CLEAN DAILY DATA
# =====================================================

tnx = tnx.reset_index()

# Keep only date and close
tnx = tnx[["Date", "Close"]]

# Yahoo returns percentages
# Example:
# 4.25 means 4.25%

tnx["rfr"] = tnx["Close"] / 100

# =====================================================
# CREATE ANNUAL AVERAGES
# =====================================================

tnx["year"] = pd.to_datetime(
    tnx["Date"]
).dt.year

rfr_annual = (
    tnx.groupby("year")["rfr"]
    .mean()
    .reset_index()
)

rfr_annual = rfr_annual.rename(
    columns={
        "year": "fiscal_year"
    }
)

# =====================================================
# SAVE FILES
# =====================================================

daily_file = os.path.join(
    base_dir,
    f"{today}_risk_free_rate_daily.csv"
)

annual_file = os.path.join(
    base_dir,
    f"{today}_risk_free_rate_annual.csv"
)

tnx.to_csv(
    daily_file,
    index=False
)

rfr_annual.to_csv(
    annual_file,
    index=False
)

print("\nAnnual Risk-Free Rates:")
print(rfr_annual)

print("\nSaved:")
print(daily_file)
print(annual_file)
```

## Phase 2: Financials building and validation

Phase 2.1: Create the Company-Year Panel

```{python}
# =====================================================
# PHASE 2.1
# BUILD CLEAN COMPANY-YEAR PANEL
# =====================================================

import pandas as pd
import os
from datetime import datetime

today = datetime.today().strftime("%Y%m%d")

# =====================================================
# LOAD COMBINED FILES
# =====================================================

income = pd.read_csv(
    os.path.join(base_dir, f"{today}_ALL_income_statements.csv")
)

balance = pd.read_csv(
    os.path.join(base_dir, f"{today}_ALL_balance_sheets.csv")
)

cashflow = pd.read_csv(
    os.path.join(base_dir, f"{today}_ALL_cashflows.csv")
)

info = pd.read_csv(
    os.path.join(base_dir, f"{today}_ALL_company_info.csv")
)

prices = pd.read_csv(
    os.path.join(base_dir, f"{today}_ALL_price_history.csv")
)

print("Files loaded.")
```

Phase 2.2: Extract Required Income Statement Fields

```{python}
# =====================================================
# EXTRACT INCOME STATEMENT VARIABLES
# =====================================================

income_fields = {
    "Total Revenue": "revenue",
    "EBIT": "ebit",
    "EBITDA": "ebitda",
    "Interest Expense": "interest_expense",
    "Net Income": "net_income"
}

income_clean = (
    income[
        income["line_item"].isin(income_fields.keys())
    ]
    .copy()
)

income_clean["variable"] = (
    income_clean["line_item"]
    .map(income_fields)
)

income_clean["fiscal_year"] = pd.to_datetime(
    income_clean["fiscal_date"]
).dt.year

income_panel = (
    income_clean
    .pivot_table(
        index=[
            "ticker",
            "company_name",
            "fiscal_year"
        ],
        columns="variable",
        values="value",
        aggfunc="first"
    )
    .reset_index()
)
income_panel.columns.name = None

print(income_panel.head())
```

Phase 2.3: Extract Balance Sheet Variables

```{python}
# =====================================================
# EXTRACT BALANCE SHEET VARIABLES
# =====================================================

balance_fields = {
    "Total Assets": "total_assets",
    "Current Assets": "current_assets",
    "Current Liabilities": "current_liabilities",
    "Working Capital": "working_capital",
    "Total Liabilities Net Minority Interest": "total_liabilities",
    "Total Debt": "total_debt",
    "Long Term Debt": "long_term_debt",
    "Current Debt": "current_debt",
    "Stockholders Equity": "equity",
    "Retained Earnings": "retained_earnings",
    "Cash And Cash Equivalents": "cash",
    "Net Debt": "net_debt"
}

balance_clean = (
    balance[
        balance["line_item"].isin(balance_fields.keys())
    ]
    .copy()
)

balance_clean["variable"] = (
    balance_clean["line_item"]
    .map(balance_fields)
)

balance_clean["fiscal_year"] = pd.to_datetime(
    balance_clean["fiscal_date"]
).dt.year

balance_panel = (
    balance_clean
    .pivot_table(
        index=[
            "ticker",
            "company_name",
            "fiscal_year"
        ],
        columns="variable",
        values="value",
        aggfunc="first"
    )
    .reset_index()
)
balance_panel.columns.name = None
print(balance_panel.head())
```

Phase 2.4: Extract Cash Flow Variables

```{python}
# =====================================================
# EXTRACT CASH FLOW VARIABLES
# =====================================================

cashflow_fields = {
    "Operating Cash Flow": "operating_cash_flow",
    "Free Cash Flow": "free_cash_flow",
    "Capital Expenditure": "capex"
}

cashflow_clean = (
    cashflow[
        cashflow["line_item"].isin(cashflow_fields.keys())
    ]
    .copy()
)

cashflow_clean["variable"] = (
    cashflow_clean["line_item"]
    .map(cashflow_fields)
)

cashflow_clean["fiscal_year"] = pd.to_datetime(
    cashflow_clean["fiscal_date"]
).dt.year

cashflow_panel = (
    cashflow_clean
    .pivot_table(
        index=[
            "ticker",
            "company_name",
            "fiscal_year"
        ],
        columns="variable",
        values="value",
        aggfunc="first"
    )
    .reset_index()
)
cashflow_panel.columns.name = None
print(cashflow_panel.head())
```

Phase 2.5: Calculate Annual Equity Volatility

```{python}
# =====================================================
# CALCULATE ANNUAL EQUITY VOLATILITY
# =====================================================

# Convert dates to UTC first
prices["Date"] = pd.to_datetime(
    prices["Date"],
    utc=True,
    errors="coerce"
)

# Remove any rows that failed conversion
prices = prices.dropna(subset=["Date"])

# Sort data
prices = prices.sort_values(
    ["ticker", "Date"]
)

# Daily returns based on Close price
prices["daily_return"] = (
    prices.groupby("ticker")["Close"]
    .pct_change()
)

# Extract year
prices["fiscal_year"] = (
    prices["Date"]
    .dt.year
)

# Annualized volatility
volatility = (
    prices.groupby(
        ["ticker", "fiscal_year"]
    )["daily_return"]
    .std()
    .reset_index()
)

volatility["equity_volatility"] = (
    volatility["daily_return"]
    * (252 ** 0.5)
)

volatility = volatility.drop(
    columns=["daily_return"]
)

print(volatility.head())
print(volatility.shape)
```

Phase 2.6: Extracting historical market cap

```{python}
# =====================================================
# HISTORICAL MARKET CAP
# =====================================================

# Extract fiscal year from prices

prices["fiscal_year"] = (
    pd.to_datetime(prices["Date"])
    .dt.year
)

# Sort dates

prices = prices.sort_values(
    ["ticker", "Date"]
)

# Keep last trading day of each year

year_end_prices = (
    prices.groupby(
        ["ticker", "fiscal_year"]
    )
    .tail(1)
)

# Get shares outstanding

shares = info[
    [
        "ticker",
        "sharesOutstanding"
    ]
].copy()

# Merge shares with year-end prices

historical_market_cap = (
    year_end_prices[
        [
            "ticker",
            "fiscal_year",
            "Close"
        ]
    ]
    .merge(
        shares,
        on="ticker",
        how="left"
    )
)

# Calculate market cap

historical_market_cap[
    "historical_market_cap"
] = (
    historical_market_cap["Close"]
    *
    historical_market_cap["sharesOutstanding"]
)

print(
    historical_market_cap.head()
)

historical_market_cap.to_csv(
    os.path.join(
        base_dir,
        f"{today}_historical_market_cap.csv"
    ),
    index=False
)
```

Phase 2.7: Build Final Company-Year Panel

```{python}
# =====================================================
# MERGE EVERYTHING
# =====================================================

panel = (
    income_panel
    .merge(
        balance_panel,
        on=[
            "ticker",
            "company_name",
            "fiscal_year"
        ],
        how="outer"
    )
    .merge(
        cashflow_panel,
        on=[
            "ticker",
            "company_name",
            "fiscal_year"
        ],
        how="outer"
    )
)

# ==========================================
# MERGE COMPANY INFO
# ==========================================

info_subset = info[[
    "ticker",
    "enterpriseValue",
    "sharesOutstanding",
    "beta"
]]

info_subset = info_subset.drop_duplicates(
    subset=["ticker"]
)

panel = panel.merge(
    info_subset,
    on="ticker",
    how="left"
)

# ==========================================
# MERGE HISTORICAL MARKET CAP
# ==========================================

panel = panel.merge(
    historical_market_cap[
        [
            "ticker",
            "fiscal_year",
            "historical_market_cap"
        ]
    ],
    on=[
        "ticker",
        "fiscal_year"
    ],
    how="left"
)

# ==========================================
# MERGE VOLATILITY
# ==========================================

panel = panel.merge(
    volatility,
    on=[
        "ticker",
        "fiscal_year"
    ],
    how="left"
)

# ==========================================
# SAVE PANEL
# ==========================================

panel_file = os.path.join(
    base_dir,
    f"{today}_company_year_panel.csv"
)

panel.to_csv(
    panel_file,
    index=False
)

print(panel_file)
print(panel.head())
```

Validation Chunk, to ensure all items are okay

```{python}
# =====================================================
# DATA QUALITY CHECK
# =====================================================

print("Panel Shape:")
print(panel.shape)

print("\nMissing Values:")

missing = (
    panel.isna()
    .sum()
    .sort_values(ascending=False)
)

print(missing)

print("\nRows with missing key variables:")

key_vars = [
    "revenue",
    "ebit",
    "interest_expense",
    "total_assets",
    "total_liabilities",
    "equity",
    "total_debt",
    "operating_cash_flow",
    "historical_market_cap",
    "equity_volatility"
]

print(
    panel[
        panel[key_vars].isna().any(axis=1)
    ][
        ["ticker", "company_name", "fiscal_year"]
    ]
)
```

```{python}
# =====================================================
# CHECK MISSING VARIABLES FOR PROBLEM COMPANIES
# =====================================================

problem_rows = panel[
    panel["ticker"].isin([
        "Ticker 1 from above output",
        "Ticker 2 from above output"
        #add additional tickers as required
    ])
]

pd.set_option("display.max_columns", None)

print(problem_rows)
```

```{python}
# =====================================================
# CLEAN DEBT VARIABLES
# =====================================================

# If total debt missing, estimate from
# current debt + long term debt

panel["total_debt"] = (
    panel["total_debt"]
    .fillna(0)
)

panel["current_debt"] = (
    panel["current_debt"]
    .fillna(0)
)

panel["long_term_debt"] = (
    panel["long_term_debt"]
    .fillna(0)
)

estimated_debt = (
    panel["current_debt"]
    + panel["long_term_debt"]
)

panel["total_debt"] = panel["total_debt"].where(
    panel["total_debt"] > 0,
    estimated_debt
)

print("Debt cleanup complete.")
```

```{python}
# =====================================================
# MERGE RISK-FREE RATES INTO PANEL
# =====================================================

rfr = pd.read_csv(
    os.path.join(
        base_dir,
        f"{today}_risk_free_rate_annual.csv"
    )
)

panel = panel.merge(
    rfr,
    on="fiscal_year",
    how="left"
)

print(panel[
    [
        "ticker",
        "fiscal_year",
        "rfr"
    ]
].head(10))

# Save updated panel

panel.to_csv(
    os.path.join(
        base_dir,
        f"{today}_company_year_panel.csv"
    ),
    index=False
)

print("Panel updated with risk-free rates.")
```

## Phase 3: Creating comparison models

First we ensure that we skip calculations for any values which have missing measures

```{python}
# =====================================================
# PHASE 3.0
# CREDIT RISK CALCULATION RULES
# =====================================================

import numpy as np

# -----------------------------------------
# Current Ratio
# Current Assets / Current Liabilities
# -----------------------------------------

panel["current_ratio_valid"] = (
    panel["current_assets"].notna()
    &
    panel["current_liabilities"].notna()
    &
    (panel["current_liabilities"] != 0)
)

# -----------------------------------------
# Debt To Equity
# Total Debt / Equity
# -----------------------------------------

panel["debt_to_equity_valid"] = (
    panel["total_debt"].notna()
    &
    panel["equity"].notna()
    &
    (panel["equity"] != 0)
)

# -----------------------------------------
# Cash Flow To Debt
# Operating CF / Total Debt
# -----------------------------------------

panel["cashflow_to_debt_valid"] = (
    panel["operating_cash_flow"].notna()
    &
    panel["total_debt"].notna()
    &
    (panel["total_debt"] > 0)
)

# -----------------------------------------
# Interest Coverage
# EBIT / Interest Expense
# -----------------------------------------

panel["interest_coverage_valid"] = (
    panel["ebit"].notna()
    &
    panel["interest_expense"].notna()
    &
    (panel["interest_expense"] != 0)
)

# -----------------------------------------
# Altman Z Score
# Require all 5 components
# -----------------------------------------

panel["altman_valid"] = (
    panel["working_capital"].notna()
    &
    panel["retained_earnings"].notna()
    &
    panel["ebit"].notna()
    &
    panel["historical_market_cap"].notna()
    &
    panel["revenue"].notna()
    &
    panel["total_assets"].notna()
    &
    panel["total_liabilities"].notna()
    &
    (panel["total_assets"] > 0)
    &
    (panel["total_liabilities"] > 0)
)

# -----------------------------------------
# Merton Model
# Require:
# Equity Value
# Debt
# Volatility
# Risk Free Rate
# -----------------------------------------

panel["merton_valid"] = (
    panel["historical_market_cap"].notna()
    &
    panel["total_debt"].notna()
    &
    panel["equity_volatility"].notna()
    &
    panel["rfr"].notna()
    &
    (panel["historical_market_cap"] > 0)
    &
    (panel["total_debt"] > 0)
    &
    (panel["equity_volatility"] > 0)
)

# -----------------------------------------
# Summary
# -----------------------------------------

validation_cols = [
    "current_ratio_valid",
    "debt_to_equity_valid",
    "cashflow_to_debt_valid",
    "interest_coverage_valid",
    "altman_valid",
    "merton_valid"
]

print(
    panel[validation_cols]
    .sum()
)
```

```{python}
# =====================================================
# PHASE 3.1
# ACCOUNTING-BASED CREDIT METRICS
# =====================================================

import numpy as np

# -----------------------------------------
# Current Ratio
# -----------------------------------------

panel["current_ratio"] = np.where(
    panel["current_ratio_valid"],
    panel["current_assets"]
    / panel["current_liabilities"],
    np.nan
)

# -----------------------------------------
# Debt to Equity
# -----------------------------------------

panel["debt_to_equity"] = np.where(
    panel["debt_to_equity_valid"],
    panel["total_debt"]
    / panel["equity"],
    np.nan
)

# -----------------------------------------
# Cash Flow to Debt
# -----------------------------------------

panel["cashflow_to_debt"] = np.where(
    panel["total_debt"] > 0,
    panel["operating_cash_flow"] / panel["total_debt"],
    np.nan
)

# -----------------------------------------
# Interest Coverage
# -----------------------------------------

panel["interest_coverage"] = np.where(
    panel["interest_expense"] > 0,
    panel["ebit"] / panel["interest_expense"],
    np.nan
)

# -----------------------------------------
# Altman Z-Score
# -----------------------------------------

panel["altman_x1"] = (
    panel["working_capital"]
    / panel["total_assets"]
)

panel["altman_x2"] = (
    panel["retained_earnings"]
    / panel["total_assets"]
)

panel["altman_x3"] = (
    panel["ebit"]
    / panel["total_assets"]
)

panel["altman_x4"] = (
    panel["historical_market_cap"]
    / panel["total_liabilities"]
)

panel["altman_x5"] = (
    panel["revenue"]
    / panel["total_assets"]
)

panel["altman_z_score"] = np.where(
    panel["altman_valid"],
    (
        1.2 * panel["altman_x1"]
        + 1.4 * panel["altman_x2"]
        + 3.3 * panel["altman_x3"]
        + 0.6 * panel["altman_x4"]
        + 1.0 * panel["altman_x5"]
    ),
    np.nan
)

# -----------------------------------------
# Preview
# -----------------------------------------

cols = [
    "ticker",
    "fiscal_year",
    "current_ratio",
    "debt_to_equity",
    "cashflow_to_debt",
    "interest_coverage",
    "altman_z_score"
]

print(panel[cols].head(20))
```

Sanity checks:

```{python}
panel[
    [
        "ticker",
        "fiscal_year",
        "altman_z_score"
    ]
].sort_values(
    "altman_z_score"
).head(10)
```

```{python}
panel[
    [
        "ticker",
        "fiscal_year",
        "altman_z_score"
    ]
].sort_values(
    "altman_z_score",
    ascending=False
).head(10)
```

Checkers of the problems above:

```{python}
panel[
    [
        "ticker",
        "fiscal_year",
        "current_ratio",
        "debt_to_equity",
        "cashflow_to_debt",
        "interest_coverage",
        "altman_z_score"
    ]
].describe()
```

Adding in Data Flags for the companies with bad data:

```{python}
# =====================================================
# DATA FLAGS
# =====================================================

panel["flag_negative_equity"] = (
    panel["equity"] < 0
)

panel["flag_negative_ebit"] = (
    panel["ebit"] < 0
)

panel["flag_negative_ocf"] = (
    panel["operating_cash_flow"] < 0
)

panel["flag_zero_debt"] = (
    panel["total_debt"] <= 0
)

panel["flag_missing_key_data"] = (
    ~panel["merton_valid"]
)

print(
    panel[
        [
            "ticker",
            "fiscal_year",
            "flag_negative_equity",
            "flag_negative_ebit",
            "flag_negative_ocf",
            "flag_zero_debt",
            "flag_missing_key_data"
        ]
    ].head(20)
)
```

Now let's implement Merton's model for each company-year:

Observed Inputs:

-   E = Market Value of Equity (marketCap)

-   σE = Equity Volatility

-   K = Total Debt

-   r = Risk-Free Rate

-   T = 1 year

We will solve for:

-   VA = Asset Value

-   σA = Asset Volatility

Using Merton's equations:

-   E = VA\*N(d1) - K\*exp(-rT)\*N(d2)

-   σE\*E = N(d1)\*σA\*VA

where:

-   d1 = \[ln(VA/K) + (r + 0.5σA²)T\] / (σA√T)

-   d2 = d1 - σA√T

Then:

1.  Distance To Default (DD)

DD = \[ln(VA/K) + (r - 0.5σA²)T\] / (σA√T)

2.  Probability of Default

PD = N(-DD)

```{python}
# =====================================================
# PHASE 3.2
# MERTON STRUCTURAL MODEL
# WITH SOLVER DIAGNOSTICS
# =====================================================

import numpy as np
import pandas as pd

from scipy.optimize import fsolve
from scipy.stats import norm

# ==========================================
# SETTINGS
# ==========================================

T = 1.0

# ==========================================
# STORAGE
# ==========================================

asset_values = []
asset_vols = []
distance_to_default = []
prob_default = []

# Solver audit fields
merton_solver_ier = []
merton_solver_message = []
merton_solver_status = []

# ==========================================
# LOOP THROUGH COMPANY-YEAR OBSERVATIONS
# ==========================================

for _, row in panel.iterrows():

    # --------------------------------------
    # Skip invalid observations
    # --------------------------------------

    if not row["merton_valid"]:

        asset_values.append(np.nan)
        asset_vols.append(np.nan)
        distance_to_default.append(np.nan)
        prob_default.append(np.nan)

        merton_solver_ier.append(np.nan)
        merton_solver_message.append(
            "Merton model not attempted because required input data was missing or invalid."
        )
        merton_solver_status.append(
            "Not attempted - invalid input data"
        )

        continue

    try:

        E = float(row["historical_market_cap"])
        sigma_E = float(row["equity_volatility"])
        K = float(row["total_debt"])
        r = float(row["rfr"])

        # ----------------------------------
        # Initial guesses
        # ----------------------------------

        VA0 = E + K
        sigma_A0 = sigma_E

        # ----------------------------------
        # System of equations
        # ----------------------------------

        def equations(x):

            VA, sigma_A = x

            d1 = (
                np.log(VA / K)
                + (r + 0.5 * sigma_A**2) * T
            ) / (
                sigma_A * np.sqrt(T)
            )

            d2 = d1 - sigma_A * np.sqrt(T)

            eq1 = (
                VA * norm.cdf(d1)
                - K * np.exp(-r * T) * norm.cdf(d2)
                - E
            )

            eq2 = (
                norm.cdf(d1)
                * sigma_A
                * VA
                - sigma_E
                * E
            )

            return [eq1, eq2]

        # ----------------------------------
        # Solve with convergence diagnostics
        # ----------------------------------

        solution, info, ier, msg = fsolve(
            equations,
            [VA0, sigma_A0],
            full_output=True
        )

        # Store raw solver outputs
        merton_solver_ier.append(ier)
        merton_solver_message.append(msg)

        # Interpret solver result
        if ier == 1:

            merton_solver_status.append(
                "Converged successfully"
            )

        else:

            merton_solver_status.append(
                "Warning: stopped before reaching a reliable solution"
            )

            asset_values.append(np.nan)
            asset_vols.append(np.nan)
            distance_to_default.append(np.nan)
            prob_default.append(np.nan)

            continue

        VA = solution[0]
        sigma_A = solution[1]

        # ----------------------------------
        # Sanity checks
        # ----------------------------------

        if (
            VA <= 0
            or sigma_A <= 0
            or np.isnan(VA)
            or np.isnan(sigma_A)
        ):

            asset_values.append(np.nan)
            asset_vols.append(np.nan)
            distance_to_default.append(np.nan)
            prob_default.append(np.nan)

            merton_solver_status[-1] = (
                "Warning: solver converged but produced invalid asset value or asset volatility"
            )

            continue

        # ----------------------------------
        # Distance To Default
        # ----------------------------------

        DD = (
            np.log(VA / K)
            + (
                r
                - 0.5 * sigma_A**2
            ) * T
        ) / (
            sigma_A * np.sqrt(T)
        )

        PD = norm.cdf(-DD)

        asset_values.append(VA)
        asset_vols.append(sigma_A)
        distance_to_default.append(DD)
        prob_default.append(PD)

    except Exception as e:

        asset_values.append(np.nan)
        asset_vols.append(np.nan)
        distance_to_default.append(np.nan)
        prob_default.append(np.nan)

        merton_solver_ier.append(np.nan)
        merton_solver_message.append(str(e))
        merton_solver_status.append(
            "Error: Merton calculation failed"
        )

# ==========================================
# SAFETY CHECK BEFORE SAVING TO PANEL
# ==========================================

print("Panel rows:", len(panel))
print("Asset values:", len(asset_values))
print("Asset vols:", len(asset_vols))
print("DD values:", len(distance_to_default))
print("PD values:", len(prob_default))
print("Solver ier values:", len(merton_solver_ier))
print("Solver messages:", len(merton_solver_message))
print("Solver statuses:", len(merton_solver_status))

# ==========================================
# SAVE RESULTS TO PANEL
# ==========================================

panel["asset_value"] = asset_values
panel["asset_volatility"] = asset_vols

panel["distance_to_default"] = distance_to_default
panel["probability_of_default"] = prob_default

panel["merton_solver_ier"] = merton_solver_ier
panel["merton_solver_message"] = merton_solver_message
panel["merton_solver_status"] = merton_solver_status

# ==========================================
# DISPLAY RESULTS
# ==========================================

print(
    panel[
        [
            "ticker",
            "fiscal_year",
            "distance_to_default",
            "probability_of_default",
            "merton_solver_ier",
            "merton_solver_status"
        ]
    ].head(20)
)

print("\nMerton solver status summary:")

print(
    panel["merton_solver_status"]
    .value_counts(dropna=False)
)
```

Data Completeness:

```{python}
# =====================================================
# DATA COMPLETENESS
# =====================================================

def assign_data_completeness(row):

    if (
        row["flag_missing_key_data"]
        or row["merton_solver_status"]
            == "Not attempted - invalid input data"
    ):
        return "Insufficient"

    warning_count = sum([
        row["flag_negative_equity"],
        row["flag_negative_ebit"],
        row["flag_negative_ocf"],
        row["flag_zero_debt"]
    ])

    if warning_count > 0:
        return "Partial"

    return "Complete"


def assign_data_completeness_reason(row):

    if (
        row["flag_missing_key_data"]
        or row["merton_solver_status"]
            == "Not attempted - invalid input data"
    ):
        return (
            "One or more required inputs were unavailable, preventing a complete risk assessment."
        )

    reasons = []

    if row["flag_negative_equity"]:
        reasons.append("negative equity")

    if row["flag_negative_ebit"]:
        reasons.append("negative EBIT")

    if row["flag_negative_ocf"]:
        reasons.append("negative operating cash flow")

    if row["flag_zero_debt"]:
        reasons.append("zero debt")

    if len(reasons) > 0:

        return (
            "Assessment was completed but the following warning indicators were detected: "
            + ", ".join(reasons)
            + "."
        )

    return (
        "All required inputs were available and the Merton solver converged successfully."
    )


panel["data_completeness"] = (
    panel.apply(
        assign_data_completeness,
        axis=1
    )
)

panel["data_completeness_reason"] = (
    panel.apply(
        assign_data_completeness_reason,
        axis=1
    )
)
```

```{python}
# =====================================================
# REPORTING VERSION OF PROBABILITY OF DEFAULT
# =====================================================

# Keep original decimal value
# Example:
# 0.003523 = 0.3523%

panel["probability_of_default_pct"] = (
    panel["probability_of_default"] * 100
)

# Optional: round for reporting
panel["probability_of_default_pct"] = (
    panel["probability_of_default_pct"]
    .round(4)
)

panel["probability_of_default_pct_display"] = (
    panel["probability_of_default_pct"]
    .map(lambda x: f"{x:.4f}%"
         if pd.notna(x)
         else "N/A")
)

# Preview

print(
    panel[
        [
            "ticker",
            "fiscal_year",
            "distance_to_default",
            "probability_of_default",
            "probability_of_default_pct",
            "probability_of_default_pct_display"
        ]
    ].head(20)
)
```

```{python}
panel[
    [
        "distance_to_default",
        "probability_of_default_pct",
        "probability_of_default_pct_display"
    ]
].describe()
```

```{python}
panel.sort_values(
    "probability_of_default_pct",
    ascending=False
)[
    [
        "ticker",
        "fiscal_year",
        "distance_to_default",
        "probability_of_default_pct",
        "probability_of_default_pct_display"
    ]
].head(20)
```

Risk level categorization

```{python}
# =====================================================
# PHASE 3.3
# CREDIT RISK INTERPRETATION FRAMEWORK
# ALTMAN + MERTON COMBINED VIEW
# =====================================================

import numpy as np
import pandas as pd

# -----------------------------------------
# Altman Assessment
# -----------------------------------------
# Classic Altman public-manufacturing interpretation:
# Z >= 2.99       Strong
# 1.81 to 2.99    Moderate
# Z < 1.81        Weak
# Missing          Insufficient Data
# -----------------------------------------

def assign_altman_assessment(z):

    if pd.isna(z):
        return "Insufficient Data"

    if z >= 2.99:
        return "Strong"

    elif z >= 1.81:
        return "Moderate"

    else:
        return "Weak"


panel["altman_assessment"] = (
    panel["altman_z_score"]
    .apply(assign_altman_assessment)
)


# -----------------------------------------
# Merton Assessment
# -----------------------------------------
# Based on Probability of Default %
# PD < 0.10%       Strong
# 0.10% to <1.00%  Moderate
# PD >= 1.00%      Weak
# Missing           Insufficient Data
# -----------------------------------------

def assign_merton_assessment(pd_pct):

    if pd.isna(pd_pct):
        return "Insufficient Data"

    if pd_pct >= 1:
        return "Weak"

    elif pd_pct >= 0.1:
        return "Moderate"

    else:
        return "Strong"


panel["merton_assessment"] = (
    panel["probability_of_default_pct"]
    .apply(assign_merton_assessment)
)


# -----------------------------------------
# Overall Indicator
# -----------------------------------------
# Low Risk:
#   Both Altman and Merton are not weak.
#
# High Risk:
#   Both Altman and Merton are weak.
#
# Mixed Signals:
#   Altman and Merton do not align.
#
# Insufficient Data:
#   Either model does not have enough data.
# -----------------------------------------

def assign_overall_indicator(row):

    altman = row["altman_assessment"]
    merton = row["merton_assessment"]

    if (
        altman == "Insufficient Data"
        or merton == "Insufficient Data"
    ):
        return "Insufficient Data"

    if (
        altman == "Weak"
        and merton == "Weak"
    ):
        return "High Risk"

    if (
        altman != "Weak"
        and merton != "Weak"
    ):
        return "Low Risk"

    return "Mixed Signals"


panel["overall_indicator"] = (
    panel.apply(
        assign_overall_indicator,
        axis=1
    )
)


# -----------------------------------------
# Commentary
# -----------------------------------------

def assign_commentary(row):

    altman = row["altman_assessment"]
    merton = row["merton_assessment"]
    solver_status = row.get(
        "merton_solver_status",
        ""
    )

    # Insufficient data case
    if (
        altman == "Insufficient Data"
        or merton == "Insufficient Data"
    ):

        return (
            "The credit risk interpretation is limited because either the "
            "Altman Z-Score or the Merton model could not be calculated with "
            "sufficient input data. Analysts should review the data quality "
            "flags and underlying financial statement inputs before drawing "
            "a conclusion."
        )

    # Both weak
    if (
        altman == "Weak"
        and merton == "Weak"
    ):

        return (
            "Both accounting-based and market-based indicators suggest elevated "
            "credit risk. The firm's financial statements show signs of weakness, "
            "and market-based measures also imply a higher probability of default. "
            "This alignment indicates a deteriorating credit profile."
        )

    # Altman weak, Merton not weak
    if (
        altman == "Weak"
        and merton != "Weak"
    ):

        return (
            "Market-based credit measures remain relatively stable, but "
            "accounting-based indicators suggest financial weakness. This may "
            "indicate that deterioration in the firm's financial statements has "
            "not yet been fully reflected in market valuations."
        )

    # Altman not weak, Merton weak
    if (
        altman != "Weak"
        and merton == "Weak"
    ):

        return (
            "Accounting-based indicators remain relatively stable, while "
            "market-based credit measures indicate elevated risk. This may "
            "suggest that investors are anticipating future challenges that are "
            "not yet fully visible in the firm's reported financial statements."
        )

    # Both not weak
    return (
        "The company's financial statements indicate a stable credit profile, "
        "and market-based credit measures imply a relatively low probability of "
        "default. Accounting and market signals are aligned and suggest low "
        "credit risk."
    )


panel["commentary"] = (
    panel.apply(
        assign_commentary,
        axis=1
    )
)


# -----------------------------------------
# Preview
# -----------------------------------------

print(
    panel[
        [
            "ticker",
            "fiscal_year",
            "altman_z_score",
            "altman_assessment",
            "probability_of_default_pct_display",
            "merton_assessment",
            "overall_indicator",
            "commentary"
        ]
    ].head(20)
)

print("\nOverall Indicator Summary:")

print(
    panel["overall_indicator"]
    .value_counts(dropna=False)
)
```

Export into csv:

```{python}
# =====================================================
# PHASE 3.4
# FINAL CREDIT RISK REPORT
# =====================================================

report_columns = [

    # Company information
    "ticker",
    "company_name",
    "fiscal_year",

    # Data quality flags

    "flag_negative_equity",
    "flag_negative_ebit",
    "flag_negative_ocf",
    "flag_zero_debt",
    "flag_missing_key_data",

    # Data completeness

    "data_completeness",
    "data_completeness_reason",

    # Accounting metrics
    "current_ratio",
    "debt_to_equity",
    "cashflow_to_debt",
    "interest_coverage",

    # Altman model
    "altman_z_score",
    "altman_assessment",

    # Market value input
    "historical_market_cap",

    # Merton model
    "distance_to_default",
    "probability_of_default",
    "probability_of_default_pct_display",

    # Merton solver diagnostics
    "merton_solver_ier",
    "merton_solver_status",
    "merton_solver_message",

    # Combined interpretation
    "merton_assessment",
    "overall_indicator",
    "commentary"
]

credit_risk_report = (
    panel[report_columns]
    .copy()
)

# Sort for readability

credit_risk_report = (
    credit_risk_report
    .sort_values(
        [
            "ticker",
            "fiscal_year"
        ]
    )
)

# ==========================================
# SAVE REPORT
# ==========================================

report_file = os.path.join(
    base_dir,
    f"{today}_credit_risk_report.csv"
)

credit_risk_report.to_csv(
    report_file,
    index=False
)

# ==========================================
# DISPLAY SUMMARY
# ==========================================

print("\nCredit Risk Report Saved:")
print(report_file)

print("\nRisk Category Summary:")

print(
    credit_risk_report["overall_indicator"]
    .value_counts(dropna=False)
)

print("\nPreview:")

print(
    credit_risk_report.head(20)
)
```

## Phase 4: Latest report creation

```{python}
# =====================================================
# LATEST COMPLETE COMPANY REPORT
# =====================================================

# A usable row is one where we have both
# accounting and market-based risk interpretations

credit_risk_report["usable_record"] = (
    credit_risk_report["altman_z_score"].notna()
    &
    credit_risk_report["distance_to_default"].notna()
    &
    credit_risk_report["probability_of_default"].notna()
    &
    (
        credit_risk_report["altman_assessment"]
        != "Insufficient Data"
    )
    &
    (
        credit_risk_report["merton_assessment"]
        != "Insufficient Data"
    )
)

# Sort newest year first

working = (
    credit_risk_report
    .sort_values(
        ["ticker", "fiscal_year"],
        ascending=[True, False]
    )
)

latest_rows = []

for ticker in working["ticker"].unique():

    company = working[
        working["ticker"] == ticker
    ]

    usable = company[
        company["usable_record"]
    ]

    # Take newest usable row
    if len(usable) > 0:

        latest_rows.append(
            usable.iloc[0]
        )

    # Otherwise take newest available row
    else:

        latest_rows.append(
            company.iloc[0]
        )

latest_company_report = pd.DataFrame(
    latest_rows
)

latest_company_report.drop(
    columns=["usable_record"],
    inplace=True,
    errors="ignore"
)

# Save CSV

latest_file = os.path.join(
    base_dir,
    f"{today}_latest_company_credit_risk_report.csv"
)

latest_company_report.to_csv(
    latest_file,
    index=False
)

print(latest_file)

print(
    latest_company_report[
        [
            "ticker",
            "company_name",
            "fiscal_year",
            "altman_z_score",
            "altman_assessment",
            "probability_of_default_pct_display",
            "merton_assessment",
            "overall_indicator",
            "commentary"
        ]
    ]
)
```
